---
title: "Linear models with categories"
format: html
editor: source
---

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
library('DT')
library(LaCroixColoR)
palette(lacroix_palette("Mango", type = "discrete"))
```

## What are we doing here?

We previously discussed linear models and learned to build a very simple linear model to estimate the mean. We also learned how to calculate residuals, which in this case were the differences between each datapoint and the mean.

In this section we'll expand our linear model to incorporate differences between categories.

## Motivating example

You may have noticed in the plot of penguin body mass in the last chapter that there appeared to be differences between different groups. Let's look more closely by making a plot of the body masses by color.

```{r}
plot(penguins$body_mass, bty="n", ylab = "body mass", 
     col=penguins$species, lwd=2)
legend('topleft',levels(penguins$species),bty="n", pch=1, pt.lwd=2, col=palette())

```

It definitely looks like something is different between these different species.
But can we use linear models to confidently say so?

## Estimating conditional means


We want to start by estimating the mean body mass of each species of penguin. 
The code below does this

```{r}
penguins                                             |>
  filter(!is.na(body_mass) , !is.na(species)) |>  #remove NAs
  group_by(species)                             |>
  summarise(mean_body_mass = mean(body_mass))
```

You may be wondering why I am referring to this as a conditional mean.
To explain this I'll walk through the math of the model, focussing only on Adelie and Gentoo penguins.

We will model body mass of an individual penguin, $i$, conditional on its species.

$$\text{Mass}_i = b_0 + b_1 \times \text{Adelie}_i + e_i $$

* $b_0$ is the intercept and, in this case, the mean body mass of Gentoo penguins
* $b_1$ is the difference in mass between Adelie and Gentoo penguins.
* $\text{Adelie}_i$ is an indicator variable that is $1$ if individual $i$ is an Adelie penguin and $0$ if individual $i$ is a Gentoo penguin.
* $e_i$ is the residual 

We could rewrite the above equation as two equations that describe the mass of individuals conditional on whether they are Adelie or Gentoo.
$$ \text{Mass}_{i|\text{Adelie}} = b_0 + e_i$$
$$ \text{Mass}_{i|\text{Gentoo}} = b_0 + b_1 + e_i$$

We can work through these equations to estimate the parameters of the model. The mean mass of an Adelie penguin ($b_0$) is 3701 and since the mean mass of a Gentoo penguin is 5076, $b_1 =$ 1375$.

## Fitting a model with lm()

```{r}
## first filter out chinstraps and anything with missing data
penguins2 <- penguins                                             |>
  filter(!is.na(body_mass) , !is.na(species), species != "Chinstrap") |> 
  droplevels()

model1 <- lm(body_mass ~ species, data=penguins2)
summary(model1)
```

## Residuals

:::fyi
**Linear models vs t-tests**

How is what we've just done different from a t-test? Or is it?

```{r}
adelies = dplyr::filter(penguins, species=="Adelie")
gentoos = dplyr::filter(penguins, species=="Gentoo")

t.test(adelies$body_mass, gentoos$body_mass)

```

Both a t-test and the linear models we've discussed so far are based on a normal distribution.
However, the t-test uses a t distribution. The t-distribution incorporates uncertainty in our estimate of the standard deviation of a normal distribution. The more data we have, the closer the t-distribution approximates a normal distribution.

Note that we can compare the output of our t.test above and the summary of the corresponding linear model and the t-value for species is the same!

:::


## More than two categories

What if we want to look at all three species? We can use the same modelling approach as before.

$$\text{Mass}_i = b_0 + b_1 \times \text{Adelie}_i +b_2 \times \text{Chinstrap} + e_i $$

* $b_0$ is the intercept and, in this case, the mean body mass of Gentoo penguins
* $b_1$ is the difference in mass between Adelie and Gentoo penguins.
* $b_2$ is the difference in mass between Chinstrap and Gentoo penguins.
* $\text{Adelie}_i$ is an indicator variable that is $1$ if individual $i$ is an Adelie penguin and $0$ if not.
* $\text{Chinstrap}_i$ is an indicator variable that is $1$ if individual $i$ is a Chinstrap penguin and $0$ if not.
* $e_i$ is the residual 

## Building a model with more than two categories with lm()

```{r}
model1 <- lm(body_mass ~ species - 1, data = penguins)
model1
```

## A note on the reference categories.

## ANOVA

There may be times where we have multiple category values but instead of knowing the effect of each category value on the outcome, we just want to know if individuals from different species have different body masses in general.
For example, if we had 100s of penguin species, we might care less about the effect of each species than how much species matter in general. This latter value would be easier to interpret.

Answering this question requires an ANOVA, which stands for "Analysis of Variance"
Formally, we are asking if individuals from different categories are drawn from the same distribution (our null hypothesis) or different distributions (our alternative hypothesis).

### Fitting an ANOVA with in R
```{r}
#install.packages('car')
library(car)

Anova(model1)
```


