---
title: "Lab 2 -- Simple Linear Models"
format: html
editor: visual
---

This is a lab to practice simple linear models with means and categories.

We will be simulating data to use in this model.


```{r}
#| label: load-packages
#| include: false
library(tidyverse)
palette <- c( "#FF5300", "#9ED80B", "#172869", "#43B629", "#8F92A1", "#1BB6AF")
```


Pretend that you're interested in studying selection on Partridge Peas (*Chamaecrista fasciculata*). You estimate fitness (fruit count) for 200 genotypes in a common garden and want to use a linear model to estimate the mean fitness.
 
$$ \text{Fitness} \sim \beta_0 + e_i$$
$$e_i \sim N(0, \sigma^2)$$
 
 
We will start by simulating data under this model using a normal distribution. Use the following parameters:
$\beta_0$ = 55 and $\sigma$=5

```{r}
set.seed(20)
myData = rnorm(n=200, mean=55, sd=5)

```
 
Make a histogram visualizing your data

```{r}
hist(myData, col=palette()[6], border="white", main="", xlab = "fitness")
```
 
 
Estimate the mean of the population using a maximum likelihood approach with a grid search. Your grid should go from 35 to 75 and have 200 points.

```{r}


```
 