---
title: "pvalue-assignment"
output: html_document
date: "2025-11-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Form groups of 3 (2-4 is ok).

Group members:


1. Discuss the following questions for each paper. While each group member shares the information about their paper, another group member should type notes about this information into the markdown document.

a) What are the main question and results of your paper?

b) Did the paper use p values to support the results?

c) What specific cut-offs (alpha values) were used? Were these explained or justified in the paper?

d) Were there other statistical techniques used to support significance of the results?

2. Collect all the p values used in the papers of every group member and, for each p value, note whether the paper thought it was "significant".

```{r}
#code for making your dataframe of p values goes here

```

3. What proportion of reported p values across your papers was significant (as defined by each paper)? 

```{r}
#code goes here

```

4. Make a histogram of your p values. Discuss this histogram and the proportion from #3 and write some notes below about your discussion and what you find interesting. Do you think that this data matches the results you'd get if you did a bunch of random statistical tests? Why or why not? 

```{r}
#histogram goes here

```

5. What other questions do you have about p values?

